---
title: "ESM 260: Applied Marine Ecology"
subtitle: "Problem Set 4"
author: "Chase Brewster"
date: "3/03/2020"
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{fontspec}
    - \setmainfont{Calibri Light}
output: 
  html_document: default
  latex_engine: xelatex
---

#Load Packages & Read in Data
```{r packages & data, warning=FALSE, message=FALSE}

library(tidyverse)
library(wesanderson)

climate <- read_csv("climate.csv")
crab <- read_csv("crab.csv")
nuclear <- read_csv("nuclear.csv")
upwell <- read_csv("upwell.csv")

```
***

#1. Climate Change

First, there is growing pressure in Congress to relax environmental regulations affecting carbon dioxide
emissions. You have been asked to provide evidence for changes in marine communities that may be due to
climatic changes, such as increases in temperature. Your have data from a coastal site in Northern California
that had been previously surveyed in 1955. During the ensuing 65 years there have been substantial changes in
the abundances of several species. To help interpret the changes, you can look at the following information for
all species: 1) the mode of development (i.e., planktonic larvae vs. non-planktonic larvae), 2) the geographic
range (i.e., does the species range occur primarily North or South of the site, or is it Cosmopolitan - both north
and south), and 3) the trophic level (i.e., predator, filter feeder, scavenger). What are the patterns of change?
Try grouping the species by categories and see if anything emerges.
```{r climate change explore}

##going to start by filtering for different qualities of the species and see if anything emerges

climate_explore <- climate %>% 
  filter(trophic == "scavenger")

##going to run a linear model to see if any species category is significant

climate_lm <- lm(density ~ development+range+trophic, data = climate)

summary(climate_lm)

```

- there does not seem to be any patterns in planktonic vs. non-planktonic

- Northern species have all decreased in density (significant)
- Southern species have all increased in density (significant)
- there is no pattern for cosmopolitan

- there does not seem to be a pattern for predators
- there does not seem to be a pattern for filter feeders
- there does not seem to be a pattern for scavengers (all but 1 increase, but they are all southern & cosmo)

```{r, climate change graph}

##Now going to make graphs for each species category

climate_graph_range <- ggplot(climate, aes(x = species, y = density)) +
  geom_col() + 
  facet_wrap(~range) +
  theme_light()

climate_graph_range

climate_graph_development <- ggplot(climate, aes(x = species, y = density)) +
  geom_col() + 
  facet_wrap(~development) +
  theme_light()

climate_graph_development

climate_graph_trophic <- ggplot(climate, aes(x = species, y = density)) +
  geom_col() + 
  facet_wrap(~trophic) +
  theme_light()

climate_graph_trophic

```
***

#2. Nuclear Power Plant

Second, El Diablo nuclear power plant is up for a renewal of its operating permit. One issue of concern is
whether the plant has been altering the marine community near its discharge point (Diablo Cove). A monitoring
program, which started ten years before the plant became operational in 2010, followed the abundance of 3
species of invertebrates at the discharge point and at two sites (Hyperion Bay & Noway Bay) that appeared to
be at least superficially similar to the discharge point. You now also have ten years of monitoring data from the
sites after the plant became operational. Are there any detectable impacts of the plant on these species? Which
site(s) did you use as a control? Why?
```{r nuclear power data}
#diablo is the operating location
#became operational in 2010
#3 species

#create datasets for each species

nuclear_a <- nuclear %>% 
  filter(species == "A") 

nuclear_b <- nuclear %>% 
  filter(species == "B")

nuclear_c <- nuclear %>% 
  filter(species == "C")

```

```{r nuclear power graph}
#Species A
nuclear_a <- nuclear_a %>% 
  group_by(location)

nuc_a_graph <- ggplot(nuclear_a, aes(x = year, y = density)) +
  geom_line(aes(color = location), size = 1) +
  theme_light() +
  scale_color_manual(values=wes_palette(n=3, name="FantasticFox1")) +
  geom_vline(xintercept = 2010) +
  geom_smooth(method = "lm", formula = y ~ x, aes(color = location), se = FALSE, size = 0.8)

nuc_a_graph

#Species B
nuclear_b <- nuclear_b %>% 
  group_by(location)

nuc_b_graph <- ggplot(nuclear_b, aes(x = year, y = density)) +
  geom_line(aes(color = location), size = 1) +
  theme_light() +
  scale_color_manual(values=wes_palette(n=3, name="FantasticFox1")) +
  geom_vline(xintercept = 2010) +
  geom_smooth(method = "lm", formula = y ~ x, aes(color = location), se = FALSE, size = 0.8)

nuc_b_graph

#Species C
nuclear_c <- nuclear_c %>% 
  group_by(location)

nuc_c_graph <- ggplot(nuclear_c, aes(x = year, y = density)) +
  geom_line(aes(color = location), size = 1) +
  theme_light() +
  scale_color_manual(values=wes_palette(n=3, name="FantasticFox1")) +
  geom_vline(xintercept = 2010) +
  geom_smooth(method = "lm", formula = y ~ x, aes(color = location), se = FALSE, size = 0.8)

nuc_c_graph
```
 
 - there is no discernable trend for pre-2010 and post-2010 for Species A
 - there is no discernable trend for pre-2010 and post-2010 for Species B
 - there appears be a slightly greater increase for Species C post-2010 relative to the control
 
 - Hyperion should serve as the control, relative to noway, as noway is highly variable and hyperion densities are closely aligned to diablo densities pre-2010

***

#3. Fisheries and Invasive Species

Finally, you have been asked to make recommendations regarding 2 key species of interest. One species is a
crab (Cancer cancer). There is a great deal of interest in developing fisheries for Cancer cancer in both
Oregon and California. The other species is a barnacle (Barnacle bill). Barnacle bill invaded the US West
Coast possibly from Japan earlier in the century. It has become a nuisance species and regularly fouls pipes,
ships etc. There is a great deal of interest in learning more about the dynamics of this species in an effort to
eliminate it.

Your job is to coordinate the new crab fishery for the West Coast of the US. In an attempt to regulate this new
fishery more effectively, you recall that variation in the recruitment rate (i.e., the # of new individuals settling
into the population) is one big problem in designing an effective management program. The traditional approach of looking for a simple relationship between the recruitment of young and the existing # of
reproductive adults has failed repeatedly, and you want to explore new approaches. You realize that the
problem lies in figuring out what factors cause the variable recruitment of young. You realize also that this
information would be useful to understand the dynamics of the barnacle in an effort to eliminate it. You request
all available data on the recruitment of both species from the past 20 years for both Northern California and
Oregon.

You quickly realize that the # of recruits varies dramatically in both species. This widespread recruitment
variation in both species is problematic. If you don't know how many new crabs will enter the population, you
cannot effectively determine the # of adults to harvest. Unless you can unravel the causes, you will not be able
to effectively manage the new fishery or understand the population dynamics of the barnacle. You posit that the
most likely limitations of prior attempts to predict recruitment solely on the basis of the # of adults are:

1) the predictions ignore the fact that the species have open populations (i.e., they have large scale larval
dispersal so that recruits may come from somewhere else),
2) the predictions ignore environmental variation affecting adults (e.g., food supply, predators), and
3) the predictions ignore environmental variation affecting the success of larvae during their development
(e.g., food supply, temperature, and predator abundances).

To rectify these limitations, you ask for all available information on the stock size of adults and on patterns of
environmental variation.

Remember: if there is a stock-recruitment relationship, the stock size (the # of adults) in one year affects
the recruitment rate in the next year (i.e., the recruits in 2004 come from adults censused in 2003).

• Cancer cancer larvae live approximately 3 weeks before settling, whereas Barnacle bill larvae live only 10
days. Both species spawn in the spring.
Key Point: At this time of year, the currents run from north to south along the shore.

• Barnacle bill are immobile and filter phytoplankton from the water. Cancer cancer are scavengers and eat
detritus that settles onto the seafloor form the overlying water.

To get information on food availability, you only have upwelling indices. Upwelling is an important coastal
process that brings nutrients up to the surface. The added nutrients generate blooms of phytoplankton. The rate
of upwelling is measured by an index that can be positive or negative. Large positive values imply strong
upwelling (and abundant nutrients and phytoplankton). Negative values imply downwelling (and low nutrients
and rare phytoplankton). You obtain values for the summer, when juveniles are developing and for the winter,
when adults are producing new eggs. Remember, values affecting adults are for different years than values
affecting larvae.

What patterns do you see in the recruitment of these species at these two sites relative to these other variables?
Try plotting the recruitment data against the other variables that might be able to explain them. Use a
scatterplot. Look for patterns (e.g., straight lines or smooth curves) that imply there is a relationship between
the variable and the pattern of recruitment. In your write-up, only include plots of the patterns that you find.
You don't need to document the patterns that are not there.
```{r}
#going to graph scatter plots comparing different variables on x and y axes

#adult vs. recruit crab california
crab_1 <- ggplot(crab, aes(x = crab_ca_rec, y = crab_ca_ad)) +
  geom_point() +
  theme_light()

crab_1
#no pattern

#adult vs. recruit crab oregon
crab_2 <- ggplot(crab, aes(x = crab_or_rec, y = crab_or_ad)) +
  geom_point() +
  theme_light() +
  geom_smooth(method = "lm", formula = y~x, se = FALSE, size = 0.5)

crab_2
#no pattern

#adult ca vs. recruit or crab
crab_3 <- ggplot(crab, aes(x = crab_or_rec, y = crab_ca_ad)) +
  geom_point() +
  theme_light()

crab_3


```



 
